Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study (draft only, not to be quoted)

### 2020-06-23 casecontrol.Rmd
parallelized stepwise variable selection
debuggedtoworkwith15Maydataset

# Tables
```{r demog, echo=FALSE, warning=FALSE, message=FALSE}
options(knitr.kable.NA = '')

knitr::kable(table.demog.append[, 1:5],
             escape=FALSE, 
             booktabs=TRUE,
             label="demog",
             row.names=FALSE,
             col.names=c("", clean.header(colnames(table.demog.aug)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align="lcccc", 
             caption="Univariate associations of severe disease with demographic factors") %>%
    column_spec(column=1, width="3cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="1.8cm") %>%
    column_spec(column=4, width="2.5cm") %>%
    kableExtra::group_rows("Ethnicity based on name classification", 1, 3, bold=FALSE) %>%
    kableExtra::group_rows("SIMD quintile", 4, 8, bold=FALSE) %>%
    kableExtra::group_rows("Ethnicity based on Scottish Morbidity Record", 10, 13, bold=FALSE,
               hline_before=TRUE, latex_gap_space="10pt") %>% 
    kable_styling(latex_options="hold_position")

@@

```{r agegr.withNRS, echo=FALSE, warning=FALSE, message=FALSE}

table.agegr.withNRS <- readRDS(table.agegr, file="table.agegr.withNRS.rds")

>>>>

# Tables
```{r demog, echo=FALSE, warning=FALSE, message=FALSE}
options(knitr.kable.NA = '')
knitr::kable(table.demog.append[, 1:5],
             escape=FALSE, 
             booktabs=TRUE,
             label="demog",
             row.names=FALSE,
             col.names=c("", clean.header(colnames(table.demog.aug)[1:2]),
                         "Rate ratio (95\\% CI)", "$p$-value"),
             align="lcccc", 
             caption="Univariate associations of severe disease with demographic factors") %>%
    column_spec(column=1, width="3cm") %>%
    column_spec(column=2, width="2cm") %>%
    column_spec(column=3, width="1.8cm") %>%
    column_spec(column=4, width="2.5cm") %>%
    kableExtra::group_rows("Ethnicity based on name classification", 1, 3, bold=FALSE) %>%
    kableExtra::group_rows("SIMD quintile", 4, 8, bold=FALSE) %>%
    kableExtra::group_rows("Ethnicity based on Scottish Morbidity Record", 10, 13, bold=FALSE,
               hline_before=TRUE, latex_gap_space="10pt") %>% 
    kable_styling(latex_options="hold_position")

@@

```{r agegr.withNRS, echo=FALSE, warning=FALSE, message=FALSE}
table.agegr.withNRS <- readRDS(table.agegr, file="table.agegr.withNRS.rds")

### 2020-06-23 pharmaco.Rmd
debuggedtoworkwith15Maydataset

*Findings* -- In those without a diagnosis of a listed condition, severe COVID-19 was strongly associated with polypharmacy, defined by total number of drug classes dispensed. With those who had no dispensed prescriptions as reference group, the rate ratio (95% CI) for COVID-19 associated with dispensing of 12 or more drug classes was `r table.numdrugsgr.listed.any[6, 4]` in those without listed conditions.  In those with listed conditions, this association was in the reverse direction: the corresponding rate ratio was `r table.numdrugsgr.listed.any[12, 4]`.  The association was specifically with non-cardiovascular drugs: in a model with number of cardiovascular drug classes and number of other drug classes as covariates, the rate ratios were respectively `r table.jointcardiovasc[1, 4]` and `r table.jointcardiovasc[2, 4]`. 
A backwards elimination procedure identified proton pump inhibitors, and analgesics (both opiates and non-opiate / compound preparations) as the drug classes with the largest independent contribution to the association of COVID-19 with polypharmacy.  Further analysis of the association with proton pump inhibitors showed a dose-response relationship that was strongest in younger age groups: the rate ratio associated with average dispensed dose of 2 or more DDDs per day was `r table.dosegr.protonpump[5, 4]` in those under 60 years and `r table.dosegr.protonpump[15, 4]` in those aged under 60 years.

*Conclusions* -- In Scotland severe COVID-19 is strongly associated with polypharmacy. This association is not easily explained by co-morbidity as it is strongest in those without relevant medical history.  Of all drug classes, analgesics and proton pump inhibitors make the largest independent contribution to this association.  As prescriptions of these drug classes are recognised indicators for overprescribing, it is difficult to distinguish between causality and confounding as explanations for this association.  This requires further investigation in other settings.  In the meantime existing advice on reducing inappropriate polypharmacy and limiting the prescribing of proton pump inhibitors should be reinforced.

@@

## Morbidity and drug prescribing
For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, all ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2, and from the national cancer registry. Diagnoses of diabetes were extracted from linkage to the national diabetes register.  British National Formulary (BNF) drug codes for all prescriptions dispensed since 1 June 2019 were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016], excluding prescriptions dispensed 15 days or less before testing positive for SARS-CoV-2. The BNF classification groups drugs by seven-digit subparagraph codes that generally correspond to drugs classed by mode of action.  For this analysis prescription codes from chapters 14 and above, comprising dressings, appliances, vaccines, anaesthesia and other preparations were grouped as "Other".  Defined daily doses (DDDs) of each of the five proton pump inhibitors in the BNF were obtained from the [DDD/ATC Index](https://www.whocc.no/atc_ddd_index/) of the WHO Collaborating Centre for Drug Statistics Methodology.  Morphine milligram equivalents (MMEs) for opiates were obtained from the [Faculty of Pain Medicine](https://fpm.ac.uk/opioids-aware-structured-approach-opioid-prescribing/dose-equivalents-and-changing-opioids) of the Royal College of Anaesthetists.  Average daily doses of proton pump inhibitors (as DDDs) and opiates (as MME) were calculated for each individual as the sum of (conversion factor $\times$ strength $\times$ quantity dispensed) divided by (observation period $\times$ DDD).

@@

## Associations of severe COVID-19 with drug classes
We next examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19 in those without listed conditions.  Supplementary Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 10 cases and were associated with severe disease at $p < 0.001$.  These COVID-19-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.  
A backward stepwise variable elimination procedure as described above was carried out to identify which of these drug classes contributed most to the association of COVID-19 with polypharmacy.  Figure \ref{fig:stepwise} shows how the log-likelihood changed as these variables were dropped from the model. Dropping the first `r firstrow.neg - 1` drug classes increased the log-likelihood of the model.  Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.  The last drug classes retained by backward selection, and thus the variables with largest independent explanatory power for the effect of polypharmacy, were proton pump, inhibitors, nonopioid /compound analgesics, and opioid analgesics.

@@

* Testing for a dose-response relationship

* Adjusting for prespecified covariates: care home residence, SIMD quintile, hospital diagnosis of a disorder of the esophagus, stomach or duodenum , ever-use of non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.

* Comparison between recent and less recent time windows of exposure.

## Associations of severe COVID-19 with analgesic use
Table \ref{tab:dosegr.opiate} shows the relationship of COVID-19 to average daily opiate dose (MMEs) over the observation period.  This showed a dose-response relationship.  With those who had not been dispensed opiates as baseline, the univariate rate ratio was XX in those with average dose daily of more than 50 mg morphine equivalent, reduced to YY on adjusting for prespecified covariates.

@@

## Associations of severe COVID-19 with proton pump inhibitor use
Table \ref{tab:dosegr.protonpump} shows that the rate ratio for COVID-19 increases with the dose of proton pump inhibitor, expressed as average DDDs/day over the period for which prescribing was observed. This relationship is strongest in younger age groups, and the effect size is reduced by about one quarter by adjusting for prespecified covariates. 
Of the five drugs in this class in the BNF, only two -- omeprazole and lansoprazole -- were widely prescribed: mean DDDs/day in controls were `r table.ppinames[3, 1]` and `r table.ppinames[2, 1]` respectively.  The effect size for unit increase in dose expressed as average DDDs/day were similar for lansoprazole [DDD 30 mg, rate ratio `r table.ppinames[2, 3]`] and omeprazole rate [DDD 20 mg, rate ratio `r table.ppinames[3, 3]`].  cy.

Table \ref{tab:timewindow.protonpump} compares the rate ratios associated with dispensing of at least one prescriptions for proton pump inhibitors in two time windows: up to 120 days before the cutoff date of 15 days before testing for SARS-CoV-2, more than 120 days before the cutoff date.  The rate ratio was not significantly higher in those whose only dispensed prescriptions were in the more recent time window than in those whose only dispensed prescriptions were in the less recent time window.

@@

# Discussion 
We have shown that severe COVID-19 is associated with polypharmacy, defined by the number of drugs classes dispensed during the period of observation.  We attempted to determine how much of this association is likely to be attributable to confounding by multimorbidity and frailty, and how much might be attributable to adverse effects of drugs.  Without more data on health status from primary care records, the confounders cannot be accurately measured.  We were able to show that the association with polypharmacy remains after excluding residents in care homes (who are at high risk of COVID-19 and more likely to have many drugs prescribed), remains after excluding those with conditions designated as conferring high risk, and is strongest in those who have had no hospital admissions in the last five years, who are unlikely to have multiple morbid conditions increasing susceptibility to severe COVID-19.  The rate ratios of more than 20 associated with dispensing of more than 10 drug classes are far larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017] and the risk ratios of 2 to 5 for severe COVID-19 that are associated with designated conditions in this population.  
The drug classes that are strongly associated with severe COVID-19 include drugs acting on the gastrointestinal and central nervous systems that are often overprescribed, and are used as indicators of overprescribing in the Scottish National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019].

Although a systematic search by backwards elimination identifies proton pump inhibitors as the drug class with the largest independent contribution to the association of COVID-19 with polypharmacy, this procedure is likely to retain the drug classes that are most commonly prescribed.  We have shown a clear dose-response relationship of dispensed average daily dose of proton pump inhibitors to risk of severe COVID-19.  Adjustment for covariates pre-specified as likely to confound this association -- disorders of the upper gastrointestinal tract, NSAIDs, antiplatelet agents and anticoagulants -- reduces the effect size only slightly. The rate ratios associated with exposure to proton pump inhibitors are highest in younger age groups, in whom comorbid conditions and frailty are less likely to confound this association.

<!---The propensity score is heavily weighted by other drug classes that are widely over-prescribed but unlikely to have direct effects on susceptibility to COVID-19: antidepressants, analgesics and nutritional supplements. ---> Exposure to proton pump inhibitors is not measured accurately in this study because the time of dispensing does not necessarily correspond to the time of use and because these drugs are also available over the counter.  In this situation, including many covariates in the study that are not directly related to the outcome but are strongly associated with the exposure under study may over-adjust the effect of that exposure towards the null.  By restricting the analysis to those with a propensity score of zero, who had no dispensed prescriptions for any of the `r nrow(coeffs.propensity)` drug classes that contribute to the propensity score, we are able to overcome this but the numbers of exposed cases and controls in this category are relatively small. One of the most powerful pharmacoepidemiologic methods for distinguishing causality from confounding is to compare adverse event rates by time windows of exposure among ever-users of the drug under study.  We did not detect any difference between the rate ratios associated with dispensing of a proton pump inhibitor in more recent and less recent time windows.  However most users of these drugs had prescriptions dispensed in both time windows, and users whose most recent prescription was in the earlier time window may have had a stock of tablets for use as required.

It is thus not possible at this stage to distinguish between confounding and causality as explanations for the association of severe COVID-19 with proton pump inhibitors.  Studies in other populations, where use of proton pump inhibitors is less strongly associated with polypharmacy and where more detailed measurements of health status are available, may be able to resolve this.  If the effect is causal, the best estimate of the unconfounded effect of proton pump inhibitors in Scotland may be the rate ratio of `r round(as.numeric(gsub(" \\(.*", "", table.protonpump.bypropensity[1, 3])), 1)` in those with a propensity score of zero.

As SARS-CoV-2 is at least partly an enteric infection [@wan_enteric_2020] and the ACE2 receptor is expressed in the intestine, it is plausible that proton pump inhibitors could increase the risk or severity of infection, either through short-term effects on gastric pH or through more persistent effects on the gut microbiome.  Uncertainties about the effect of proton pump inhibitors on the risk of community acquired pneumonia [@othman_community_2016] are not necessarily relevant to effects on enteric infection.  Proton pump inhibitors have been reported to affect the gut microbiome [@imhann_proton_2016].  If effects of proton pump inhibitor use on susceptiblity to COVID-19 are mediated through alteration of the gut microbiome, these effects would not necessarily be detected in a comparison of time windows. 
 <!---barrier to viral infection, or through altering the microbiome allowing with Warnings been issued about increased risk of _Cl. difficile_ infection from use proton pump inhibitors.  [ https://bnf.nice.org.uk/drug/omeprazole.html#importantSafetyInformations ] Given this possible association with increased risk of enteric infections --->.

In Scotland and other countries proton pump inhibitors are widely prescribed among the elderly, often without any clear indication [@jarchow-macdonald_prescribing_2013; @marks_time_2016; @kurdi_ongoing_2018].  Initiatives to improve prescribing quality have attempted to reduce this.  It would be of interest to examine the association between severe COVID-19 and use of proton pump inhibitors in other countries where COVID-19 epidemics have been especially severe and overprescribing of proton pump inhibitors has been reported previously [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010].

We have briefly reported associations of COVID-19 with dispensed prescriptions of drugs acting on other systems of interest. We emphasize that because of the relationship of COVID-19 to polypharmacy, associations with specific drug classes cannot be studied without taking into account how those drug classes are related to the profile of all prescribed drugs. For ACE inhibitors and angiotensin-II receptor blockers, our results are consistent with other studies that have found no increased risk associated with these drugs [@mancia_reninangiotensinaldosterone_2020].  The relation of anticoagulant use to COVID-19 is of interest because coagulopathy is a feature of severe disease [@connors_covid-19_2020].  Although in this study anticoagulants were associated with increased risk in univariate analysis, this association was was markedly reduced by adjusting for covariates including diagnosed heart disease and co-prescribing of proton pump inhibitors.  The numbers of cases and controls exposed to hydroxychloroquine in this dataset were too small for associations of severe COVID-19 with this drug to be estimated reliably.

## Methodological strengths and weaknesses
A strength of this study is that it is based on dispensed rather than issued prescriptions.  Our study does not capture all exposure to drugs that are available over the counter, which include proton pump inhibitors.  Although we do not have morbidity data from primary care which would include risk factors such as smoking and obesity, we have diagnoses from hospital discharge records that have been coded to the ICD-10 by trained coders.

@@

```{r stepwise, echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, out.width = "0.99\\textwidth", fig.cap="Stepwise dropping of drug classes from the polypharmacy variable: each row shows the coefficient of the polypharmacy variable with that variable retained, and the log-likelihood when that drug class is dropped"}

@@

knitr::kable(table.dosegr.opiate,
             escape=FALSE,
             label="dosegr.opiate", 
             booktabs=TRUE,
             row.names=TRUE, 
             col.names=c(clean.header(colnames(table.dosegr.opiate)[1:2]), 
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Associations of severe COVID-19 with dispensed average daily dose (MME) of opiate grouped into categoriesb") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Multivariate"=2)) %>%
    kableExtra::group_rows("Average dispensed daily dose (MME) of opiates", 1, 4) %>%
        kable_styling(latex_options="hold_position") %>%
        landscape()

@@

knitr::kable(table.timewindow.opiate,
             escape=FALSE,
             label="timewindow.opiate", 
             booktabs=TRUE,
             col.names=c(clean.header(colnames(table.timewindow.opiate)[1:2]), 
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption=paste0("Associations of severe disease with dispensed prescriptions for opiate analgesics by ", TW, "-day time window")) %>%
	column_spec(column=1, width="3cm") %>%
	column_spec(column=2:3, width="2cm") %>%
	column_spec(column=4, width="2.5cm") %>%
	column_spec(column=5, width="1.5cm") %>%
    kable_styling(latex_options="hold_position")

@@

knitr::kable(table.dosegr.protonpump,
             escape=FALSE,
             label="dosegr.protonpump", 
             booktabs=TRUE,
             row.names=FALSE, # dose levels in column 1
             col.names=c("Average DDDs/day", colnames(table.dosegr.protonpump)[2:3], 
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Associations of severe COVID-19 with dispensed average daily dose of proton pump inhibitor (DDDs/day) grouped into categories, with and without adjustment for prespecified covariates") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
    kable_styling(latex_options="hold_position") %>%
    kableExtra::group_rows("Ages 0-59 years", 1, 5) %>%
    kableExtra::group_rows("Ages 60-74 years", 6, 10) %>%
    kableExtra::group_rows("Ages 75+ years", 11, 15) %>%
    landscape()

>>>>

*Findings* -- In those without a diagnosis of a listed condition, severe COVID-19 was strongly associated with polypharmacy; the rate ratio (95% CI) associated with dispensing of 12 or more drug classes versus no dispensed prescriptions was `r table.numdrugsgr.listed.any[6, 4]`. In those with listed conditions, this association was in the reverse direction: the corresponding rate ratio was `r table.numdrugsgr.listed.any[12, 4]`.  The association was specifically with non-cardiovascular drugs: in a model with number of cardiovascular drug classes and number of other drug classes as covariates, the rate ratios were respectively `r table.jointcardiovasc[1, 4]` and `r table.jointcardiovasc[2, 4]`. 
A backwards elimination procedure identified proton pump inhibitors and opiate analgesics as the drug classes with the largest independent contribution to the association of COVID-19 with polypharmacy.  Further analysis of these associations showed dose-response relationships that were strongest in younger age groups and reduced only slightly by adjusting for prespecified covariates.  Stratification by time window of exposure showed that the rate ratios were higher for those with prescriptions dispensed only in the last 120 days than for those with prescriptions dispensed only in an earlier time window.

*Conclusions* -- In Scotland severe COVID-19 is strongly associated with polypharmacy. This association is not easily explained by co-morbidity as it is strongest in those without relevant medical history.  Of all drug classes, analgesics and proton pump inhibitors make the largest independent contribution to this association.  As prescriptions of these drug classes are recognised indicators for overprescribing, it is difficult to distinguish between causality and confounding as explanations for this association.  This requires further investigation in other settings.  In the meantime existing advice on reducing inappropriate polypharmacy and limiting the prescribing of opiates and proton pump inhibitors should be reinforced.

@@

## Morbidity and drug prescribing
For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, all ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2, and from the national cancer registry. Diagnoses of diabetes were extracted from linkage to the national diabetes register.  British National Formulary (BNF) drug codes for all prescriptions dispensed since 1 June 2019 were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016], excluding prescriptions dispensed 15 days or less before testing positive for SARS-CoV-2. The BNF classification groups drugs by seven-digit subparagraph codes that generally correspond to drugs classed by mode of action.  For this analysis prescription codes from chapters 14 and above, comprising dressings, appliances, vaccines, anaesthesia and other preparations were grouped as "Other".  Defined daily doses (DDDs) of each of the five proton pump inhibitors in the BNF were obtained from the [DDD/ATC Index](https://www.whocc.no/atc_ddd_index/) of the WHO Collaborating Centre for Drug Statistics Methodology.  Morphine milligram equivalents (MMEs) for opiates were obtained from the [Faculty of Pain Medicine](https://fpm.ac.uk/opioids-aware-structured-approach-opioid-prescribing/dose-equivalents-and-changing-opioids) of the Royal College of Anaesthetists.  Average daily doses of proton pump inhibitors (as DDDs) and opiates (as MME) were calculated for each individual as the sum of (conversion factor $\times$ strength $\times$ quantity dispensed) divided by (observation period $\times$ DDD).  The dispensed daily dose of opiate was calculated as the sum over all opiate-containing items, including compound analgesics.

@@

## Associations of severe COVID-19 with drug classes
We next examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19 in those without listed conditions.  Supplementary Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 10 cases and were associated with severe disease at $p < 0.001$.  These COVID-19-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.  
A backward stepwise variable elimination procedure as described above was carried out to identify which of these drug classes contributed most to the association of COVID-19 with polypharmacy.  Figure \ref{fig:stepwise} shows how the log-likelihood changed as these variables were dropped from the model. Dropping the first `r firstrow.neg - 1` drug classes increased the log-likelihood of the model.  Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.  The last drug classes retained by backward selection, and thus the variables with largest independent explanatory power for the effect of polypharmacy, were proton pump inhibitors and nonopioid or compound analgesics.

@@

* Testing for a dose-response relationship and stratifying by age group

* Adjusting for prespecified covariates: for proton pump inhibitors these covariates were specified as hospital diagnosis of a disorder of the esophagus, stomach or duodenum , ever-use of non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.  For opiate use any history of neoplasm was included as a covariate.  Care home residence and SIMD quintile were included as covariates for both these drug classes.

* Comparison between recent (last 120 days of observation) and less recent (1 June 2019 to 121 days before exposure) time windows of dispensing

## Associations of severe COVID-19 with opiate use
Table \ref{tab:dosegr.opiate} shows the relationship of COVID-19 to average daily opiate dose (MMEs) over the observation period.  This showed a dose-response relationship.  With those who had not been dispensed opiates as baseline, the univariate rate ratio was XX in those with average dose daily of more than 50 mg morphine equivalent, reduced to YY on adjusting for prespecified covariates.

@@

## Associations of severe COVID-19 with proton pump inhibitor use
Table \ref{tab:dosegr.protonpump} shows that the rate ratio for COVID-19 increases with the dose of proton pump inhibitor, expressed as average DDDs/day over the period for which prescribing was observed. This relationship is strongest in younger age groups, and the effect size is reduced by about one quarter by adjusting for prespecified covariates. 
Of the five drugs in this class in the BNF, only two -- omeprazole and lansoprazole -- were widely prescribed: mean DDDs/day in controls were `r table.ppinames[3, 1]` and `r table.ppinames[2, 1]` respectively.  The effect size for unit increase in dose expressed as average DDDs/day were similar for lansoprazole [DDD 30 mg, rate ratio `r table.ppinames[2, 3]`] and omeprazole rate [DDD 20 mg, rate ratio `r table.ppinames[3, 3]`].

Table \ref{tab:timewindow.protonpump} compares the rate ratios associated with dispensing of at least one prescriptions for proton pump inhibitors in two time windows: up to 120 days before the cutoff date of 15 days before testing for SARS-CoV-2, more than 120 days before the cutoff date.  The rate ratio was significantly higher in those whose only dispensed prescriptions were in the more recent time window than in those whose only dispensed prescriptions were in the less recent time window.

@@

# Discussion 
We have shown that severe COVID-19 is associated with polypharmacy, defined by the number of drugs classes dispensed during the period of observation.  We attempted to determine how much of this association is likely to be attributable to confounding by multimorbidity and frailty, and how much might be attributable to adverse effects of drugs.  Without more data on health status from primary care records, the confounders cannot be accurately measured.  We were able to show that the association with polypharmacy remains after excluding residents in care homes (who are at high risk of COVID-19 and more likely to have many drugs prescribed), remains after excluding those with conditions designated as conferring high risk, and is strongest in those who have had no hospital admissions in the last five years, who are unlikely to have multiple morbid conditions increasing susceptibility to severe COVID-19.  The rate ratios of more than 20 associated with dispensing of more than 10 drug classes are far larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017] and the risk ratios of 2 to 5 for severe COVID-19 that are associated with designated conditions in this population.

The drug classes that are strongly associated with severe COVID-19 include drugs acting on the gastrointestinal and central nervous systems that are often overprescribed, and are used as indicators of overprescribing in the Scottish National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019]. These indicators of overprescribing include proton pump inhibitors and analgesics, identifed by backwards elimination as the drug classes with the largest independent contribution to the association of COVID-19 with polypharmacy. 
Distinguishing between causality and confounding as possible explanations for the association of COVID-19 with these drug classes is difficult because drug exposures are not measured accurately, and the associations are likely to be confounded by other drugs, by co-morbidities and more generally by frailty and socioeconomic deprivation. The time of exposure to a drug does not necessarily correspond to the time intervals for which the prescription was issued.   Because proton pump inhibitors are available over the counter, dispensed prescriptions do not capture all exposure.

For both opiates and proton pump inhibitors, four criteria provide modest evidence favouring causal explanations over confounding.  There are strong dose-response relationships of COVID-19 to dispensed average daily dose.  Adjustment for covariates pre-specified as likely to confound these associations reduces the effect size only slightly. The effect sizes are larger in younger individuals, in whom the prevalence of comorbid conditions and frailty is lower than in older individuals.  Among ever-exposed individuals, the rate ratios associated with dispensing only in the most recent time window were higher than the rate ratios associated with dispensing only in an earlier time window.  We emphasize however that this   evidence is not conclusive. It may be possible to resolve this through studies in other countries where COVID-19 epidemics have been especially severe and overprescribing of proton pump inhibitors [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010] or opiates has been reported previously.  In Scotland proton pump inhibitors are widely prescribed among the elderly, often without any clear indication [@jarchow-macdonald_prescribing_2013; @marks_time_2016; @kurdi_ongoing_2018], but overprescribing of opiates is still relatively uncommon.

The dose-response relationship of opiate use to COVID-19 is similar in magnitude to that reported for community-acquired pneumonia in a study of people receiving medical care through the Veterans Administration before 2012 [@edelman_association_2019].  As SARS-CoV-2 is at least partly an enteric infection [@wan_enteric_2020] and the ACE2 receptor is expressed in the intestine, it is plausible that proton pump inhibitors could increase the risk or severity of infection.

We have briefly reported associations of COVID-19 with dispensed prescriptions of drugs acting on other systems of interest. We emphasize that because of the relationship of COVID-19 to polypharmacy, associations with specific drug classes cannot be studied without taking into account how those drug classes are related to the profile of drug prescribing. For ACE inhibitors and angiotensin-II receptor blockers, our results are consistent with other studies that have found no increased risk associated with these drugs [@mancia_reninangiotensinaldosterone_2020].  The relation of anticoagulant use to COVID-19 is of interest because coagulopathy is a feature of severe disease [@connors_covid-19_2020].  Although in this study anticoagulants were associated with increased risk in univariate analysis, this association was was markedly reduced by adjusting for covariates including diagnosed heart disease and co-prescribing of proton pump inhibitors.  The numbers of cases and controls exposed to hydroxychloroquine in this dataset were too small for associations of severe COVID-19 with this drug to be estimated reliably.

## Methodological strengths and weaknesses
Although we do not have morbidity data from primary care, which would include risk factors such as smoking and coding of presenting complaints, we have diagnoses from hospital discharge records that have been coded to ICD-10 by trained coders. A strength of this study is that it is based on dispensed rather than issued prescriptions.

@@

```{r stepwise, echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, out.width = "0.99\\textwidth", fig.cap="Stepwise dropping of drug classes from the polypharmacy variable: each row shows the log-likelihood when that drug class is dropped from a model with care home as covariate"}

@@

knitr::kable(table.dosegr.opiate,
             escape=FALSE,
             label="dosegr.opiate", 
             booktabs=TRUE,
             row.names=FALSE, # dose levels in column 1
             col.names=c("Average MME/day", colnames(table.dosegr.opiate)[2:3], 
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Associations of severe COVID-19 with dispensed average daily dose of opiate (MME) grouped into categories, with and without adjustment for prespecified covariates") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
    kable_styling(latex_options="hold_position") %>%
    kableExtra::group_rows("Ages 0-74 years", 1, 4) %>%
    kableExtra::group_rows("Ages 75+ years", 5, 8) %>%
    landscape()

@@

knitr::kable(table.timewindow.opiate,
             escape=FALSE,
             label="timewindow.opiate", 
             booktabs=TRUE,
             col.names=c(clean.header(colnames(table.timewindow.opiate)[1:2]), 
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption=paste0("Associations of severe disease with dispensed prescriptions for opiate-containing analgesics by ", TW, "-day time window, in those without a history of neoplasm")) %>%
	column_spec(column=1, width="3cm") %>%
	column_spec(column=2:3, width="2cm") %>%
	column_spec(column=4, width="2.5cm") %>%
	column_spec(column=5, width="1.5cm") %>%
    kable_styling(latex_options="hold_position")

@@

knitr::kable(table.dosegr.protonpump,
             escape=FALSE,
             label="dosegr.protonpump", 
             booktabs=TRUE,
             row.names=FALSE, # dose levels in column 1
             col.names=c("Average DDDs/day", colnames(table.dosegr.protonpump)[2:3], 
                         "Rate ratio (95\\% CI)", "$p$-value",
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Associations of severe COVID-19 with dispensed average daily dose of proton pump inhibitor (DDDs) grouped into categories, with and without adjustment for prespecified covariates") %>%
    add_header_above(c(" "=3, "Univariate"=2, "Adjusted for covariates"=2)) %>%
    kable_styling(latex_options="hold_position") %>%
    kableExtra::group_rows("Ages 0-74 years", 1, 5) %>%
    kableExtra::group_rows("Ages 75+ years", 6, 10) %>%
    landscape()
